---
title: "NY restaurant inspections"
author: "Ho Fai Wong"
date: "April 23, 2016"
output: 
  ioslides_presentation:
    fig_width: 7
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

#Load libraries
library(plyr)
library(dplyr)
library(zipcode)
library(ggplot2)
library(choroplethrZip)
library(lubridate)

#Load inspection data
raw = read.csv("./DOHMH_New_York_City_Restaurant_Inspection_Results.csv", stringsAsFactors = F)
raw.df = tbl_df(raw)

#Rename and re-format columns
names(raw.df) = tolower(names(raw.df))
raw.df = rename(raw.df, cuisine = cuisine.description) %>%
  mutate(., inspection.date = as.Date(inspection.date, "%m/%d/%Y")) %>%
  mutate(., grade.date = as.Date(grade.date, "%m/%d/%Y")) %>%
  mutate(., record.date = as.Date(record.date, "%m/%d/%Y")) %>%
  mutate(., phone = as.double(phone)) %>%
  mutate(., boro = factor(boro, levels(factor(boro))[c(1:3,5:6,4)])) %>%
  mutate(., zipcode = clean.zipcodes(zipcode)) %>% #Turn zip to string with clean format e.g. preceding 0s, for zip_choroplethr

  #Shorten/reword action and cuisine
  mutate(., cuisine = gsub(pattern = 'Latin \\(Cuban, Dominican, Puerto Rican, South \\& Central American\\)', replacement = 'Latin', x = cuisine, ignore.case = F)) %>%
  mutate(., cuisine = gsub(pattern = 'Café/Coffee/Tea', replacement = 'Cafe/Coffee/Tea', x = cuisine, ignore.case = F)) %>%
  mutate(., cuisine = factor(cuisine, levels(factor(cuisine))[c(1:55,57,59:84,58,56)])) %>% #Moving N/A and Other to the bottom of cuisine factors
  mutate(., action = gsub(pattern = "Violations were cited in the following area\\(s\\).", replacement = "violations", x = action, ignore.case = F)) %>%
  mutate(., action = gsub(pattern = "No violations were recorded at the time of this inspection.", replacement = "no violations", x = action, ignore.case = F)) %>%
  mutate(., action = gsub(pattern = "Establishment re-opened by DOHMH", replacement = "reopened", x = action, ignore.case = F)) %>%
  mutate(., action = gsub(pattern = "Establishment Closed by DOHMH.  violationsand those requiring immediate action were addressed.", replacement = "closed", x = action, ignore.case = F)) %>%
  mutate(., action = gsub(pattern = "Establishment re-closed by DOHMH", replacement = "reclosed", x = action, ignore.case = F)) %>%

  #Add columns for analysis
  mutate(., yearmon = as.Date(paste("1",
                                    strftime(inspection.date, "%m"), 
                                    strftime(inspection.date, "%Y"), 
                                    sep="."), 
                              format="%d.%m.%Y")) %>% #Add column for month and year of inspection
  mutate(., new_grade = ifelse(score < 0, 'Negative', 
                               ifelse(score < 14 , 'A',
                                      ifelse(score < 28, 'B', 'C')))) #Assign grades based on scores (existing data may have score but no grade)

  #Fix bad data
levels(raw.df$boro) = c('Bronx','Brooklyn','Manhattan','Queens','Staten Island','Missing')
raw.df$boro[raw.df$zipcode==11249] = 'Brooklyn' #Fill missing data...
raw.df = filter(raw.df, 
                  zipcode!='07005' & #Eliminate NJ inspection...
                  boro != 'Missing' & #Eliminate remaining rows with boro == 'Missing'; confirmed that none are in NYC
                  inspection.date > '1900-01-01' & #Eliminate rows where the restaurant hasn't been inspected yet
                  !is.na(score) & #Eliminate rows without a score
                  score >= 0 #Eliminate rows with a negative score
)


#Unique inspections
inspections = unique(select(raw.df, camis, boro, zipcode, cuisine, 
                            inspection.date, action, score, new_grade, yearmon))

#Unique inspections but only the latest for each restaurant
latest = merge(
  aggregate(inspection.date ~ camis, inspections, max), 
  inspections)

#Unique restaurants and their latest score/grade
restaurants = unique(select(latest, camis, boro, zipcode, cuisine, score, new_grade))
```

## Overview

- How does a restaurant's location, cuisine type or timing of inspection affect it's health inspection results? E.g:
- Are restaurants in Manhattan cleaner than those in Queens or the Bronx?
- Do Chinese/Asian restaurants have poorer hygiene than French or American restaurants?
- Are summer inspections associated with worse health results?

## Grades by borough

```{r}
#Bar plot: Restaurants by borough and latest grade
ggplot(data=latest, aes(x=reorder(boro, boro, function(x)-length(x)))) + 
  geom_bar(aes(fill=new_grade), position='dodge') +
  labs(title='Restaurants by borough and latest grade', 
       x='Borough', 
       y='Restaurants') +
  scale_y_continuous(limits=c(0,9000), breaks=seq(0,10000,1000)) +
  scale_fill_brewer(name="Grade", palette='Set1') +
  theme_bw() +
  theme(legend.key=element_blank())
```

## Scores by borough

```{r}
#Density: Restaurants by latest score and borough
ggplot(data=latest, aes (x=score)) + 
  stat_density(aes(color=boro), geom="line",position="identity") + 
  coord_cartesian(xlim=c(0,40)) +
  labs(title='Density of restaurants by latest score and borough', 
       x='Score', 
       y='Restaurant density') + 
  scale_y_continuous(breaks=seq(0,0.14,0.02),
                     labels = scales::percent) +
  scale_colour_brewer(name="Borough", palette='Set1') +
  theme_bw() +
  theme(legend.key=element_blank()) +
  geom_vline(xintercept = c(14,28), colour='grey') +
  annotate("text", x = c(6,20,35), y = 0.11, label = c('A','B','C'), size=6) +
  annotate("rect", xmin = 0, xmax = 14, ymin = 0, ymax = 0.13, alpha = .2, fill='darkred') +
  annotate("rect", xmin = 14, xmax = 28, ymin = 0, ymax = 0.13, alpha = .2, fill='darkblue') +
  annotate("rect", xmin = 28, xmax = 60, ymin = 0, ymax = 0.13, alpha = .2, fill='darkgreen')
```

## Average score by zipcode

```{r, warning = FALSE, message = FALSE}
#Map average scores by zipcode
avgbyzip = latest %>% 
  group_by(., zipcode) %>%
  summarise(., value=mean(score)) %>%
  rename(., region=zipcode)

#Choroplethr doesn't have all zipcodes in the NY inspection data (uses 2010 census ZCTAs)
scored_zips = unique(avgbyzip$region) #zipcodes that appear in inspection data data
missing_zips = c('10048','10055', '10057', '10104', '10105', '10106', '10107', '10118', '10121', '10123', '10155', '10166', '10175', '10176', '10178', '10179', '10281', '10285', '10317', '11242', '11249', '11256', '11352') #zipcodes that are not in choropleth package
adjusted_zips = scored_zips[!(scored_zips %in% missing_zips)] #zipcodes that appear in original data and in choropleth package

#Map: Average scores by zipcode; zipcodes are binned into 9 buckets of equal size due to limitations of the function
zip_choropleth(avgbyzip, 
               zip_zoom = adjusted_zips,
               title="Average of latest scores by zipcode",
               num_colors=5) + 
  scale_fill_brewer(palette='OrRd', name='Average score')
```


## Closures by borough

```{r}
#Total counts of inspections that led to closures by borough
inspClosByBoro = inspections %>%
  group_by(., boro) %>%
  summarise(., 
            closures = sum(action == 'closed' | action == 'reclosed'),
            inspections = n(),
            ratio = closures / inspections
  )

#Restaurants that were closed at least once (closures, not reclosures as part of the same inspection cycle)
reclosedByBoro =  inspections %>% 
  filter(., action == 'closed') %>%
  group_by(., camis, boro, zipcode) %>%
  summarize(., count = n()) %>%
  group_by(., boro) %>%
  summarize(., total_closed = n(), 
            closed_more_than_once = sum(count>1), 
            ratio = closed_more_than_once / total_closed)

#Closure and repeat closure combined
boroClosureRatios = left_join(inspClosByBoro, reclosedByBoro, by = 'boro') %>%
  rename(., inspRatio = ratio.x, recloRatio = ratio.y)

ggplot(data = boroClosureRatios, aes(x=inspRatio, y=recloRatio)) + 
  geom_point(aes(colour = boro), size = 3) +
  labs(title='Closure and repeat closure ratios by borough',
       x='Inspection closure ratio',
       y='Repeat closure ratio') +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  scale_colour_brewer(palette = 'Set1', name = 'Borough') + 
  theme_bw() +
  theme(legend.key=element_blank())
```


## Current state by cuisine

```{r}
#Down-select top 20 cuisine types for further analysis (out of 84)
restByCuisineFiltered = restaurants %>%
  group_by(., cuisine) %>%
  summarise(., count=n()) %>%
  arrange(., desc(count)) %>%
  top_n(., 20, count)

restaurantsFiltered = tbl_df(semi_join(restaurants, restByCuisineFiltered, by = 'cuisine')) #Will keep top 20 covering ~81%
inspectionsFiltered = tbl_df(semi_join(inspections, restByCuisineFiltered, by = 'cuisine'))

#Ratio of inspection closures by top 20 cuisine types
inspClosByCuisine = inspectionsFiltered %>%
  group_by(., cuisine) %>%
  summarise(., 
            closures = sum(action == 'closed' | action == 'reclosed'),
            inspections = n(),
            ratio = closures / inspections
  )

#Bar plot: Inspection closure ratio by top 20 cuisine types
ggplot(data=inspClosByCuisine, aes(x=reorder(cuisine, ratio), y=ratio)) + 
  geom_bar(stat='identity') +
  labs(title='Ratio of inspection closures by top 20 cuisine types', 
       x='Top 20 cuisine types', 
       y='Inspection closure ratio') +
  scale_y_continuous(limits=c(0,0.03),
                     breaks=seq(0,3,0.005),
                     labels = scales::percent) +
  theme_bw() +
  coord_flip()

```


## Repeat closure by cuisine

```{r}
#Restaurants that were closed at least once (closures, not reclosures as part of the same inspection cycle)
reclosedByCuisine = unique(
  inspectionsFiltered %>%
    filter(., action == 'closed') %>%
    select(., camis, cuisine, inspection.date)) %>%
  group_by(., camis, cuisine) %>%
  summarise(., closures = n()) %>% 
  group_by(., cuisine) %>%
  summarise(., closed_once = sum(closures==1), 
            closed_more_than_once = sum(closures>1), 
            total_closed = n(),
            ratio = closed_more_than_once / total_closed) 


#Bar plot: Ratio of restaurants by top 20 cuisine types that were closed more than once, out of restaurants that were closed at least once
ggplot(data = reclosedByCuisine, aes (x = reorder(cuisine, ratio), y = ratio)) + 
  geom_bar(stat='identity') +
  labs(title='Ratio of repeat closures by top 20 cuisine types', 
       x='Top 20 cuisine types', 
       y='Repeat closure ratio') +
  scale_y_continuous(limits=c(0,0.2),
                     breaks=seq(0,0.2,0.05),
                     labels=scales::percent) +
  theme_bw() +
  coord_flip()

```


## Conclusion

- Summary
- Next steps

# Appendix

## Closures by cuisine and borough

```{r}
#Ratio of inspection closures by top 20 cuisine and boro
inspClosByCuisineBoro = inspectionsFiltered %>%
  group_by(., cuisine, boro) %>%
  summarise(., 
            closures = sum(action == 'closed' | action == 'reclosed'),
            inspections = n(),
            ratio = closures / inspections
  )

#Refactor cuisines to keep same order as in previous plot
inspClosByCuisineBoro$cuisine = factor(inspClosByCuisineBoro$cuisine)
newFactor = levels(inspClosByCuisineBoro$cuisine)[order(-rank(inspClosByCuisine$ratio))]
inspClosByCuisineBoro$cuisine = factor(inspClosByCuisineBoro$cuisine, levels=rev(newFactor))

#Faceted bar plot: Inspection closure ratio by top 20 cuisine and borough
ggplot(data=inspClosByCuisineBoro, aes(x=cuisine, y=ratio)) + 
  geom_bar(stat='identity', aes(fill=boro)) + 
  facet_grid(. ~ boro) +
  labs(title='Inspection closure ratio by top 20 cuisine and borough', x='Cuisine (top 20)', y='Inspection closure ratio') +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_brewer(palette = 'Set1', name = 'Borough') +
  theme_bw() +
  theme(legend.key=element_blank(), legend.position="bottom") +
  coord_flip()

```


## Score by month

```{r}
#Trend of average scores by month and borough
trendScore = inspections %>%
  mutate(., month=month(inspection.date)) %>%
  group_by(., month, boro) %>%
  summarise(., avg = mean(score))

ggplot(data=trendScore, aes(x = month, y = avg)) + 
  geom_freqpoly(stat='identity', aes(color=boro)) +
  labs(title='Average score by month and borough',
       x='Month',
       y='Average score') +
  coord_cartesian(xlim = c(1,12), ylim = c(13,18)) +
  scale_x_continuous(breaks = 1:12, 
                     labels=c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec")) +
  scale_color_brewer(name='Borough', palette='Set1') +
  theme_bw() +
  theme(legend.key=element_blank())

```


## Closure ratio by month

```{r}
#Trend of inspection closure ratio by month and borough
trendClosures = inspections %>% 
  mutate(., month=month(inspection.date)) %>%
  group_by(., month, boro) %>%
  summarise(., 
            closures = sum(action == 'closed' | action == 'reclosed'),
            inspections = n(),
            ratio = closures / inspections
  )

ggplot(data=trendClosures, aes(x=month, y=ratio)) + 
  geom_freqpoly(stat='identity', aes(color=boro)) + 
  labs(title='Inspection closure ratio by month and borough',
       x='Month',
       y='Inspection closure ratio') +
  coord_cartesian(xlim = c(1,12)) +
  scale_x_continuous(breaks = 1:12, 
                     labels=c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec")) +
  scale_y_continuous(limits=c(0.005,0.035),
                     breaks=seq(0,0.035,0.005),
                     labels=scales::percent) +
  scale_color_brewer(name='Borough', palette='Set1') +
  theme_bw() +
  theme(legend.key=element_blank())
```
